---
title: "Data Output Lab"
output: html_document
editor_options: 
  chunk_output_type: console
---

In this lab you can use the interactive console to explore or **Knit** the document. 
Remember anything you type here can be "sent" to the console with Cmd-Enter (OS-X) or Ctrl-Enter (Windows/Linux) in an R code chunk.


# Part 1

### 1.1
Load the tidyverse package. Then read in the CalEnviroScreen dataset from 
https://daseh.org/data/CalEnviroScreen_data.csv
and assign it to an object named `ces. 

```
# General format
library(readr)
# OBJECT <- read_csv(FILE)
```

```{r 1.1response}

```

### 1.2

Filter the dataset so that the `CaliforniaCounty` column is equal to "Yuba". Store the modified dataset as `ces_Yuba`.

```
# General format
NEW_OBJECT <- OBJECT %>% filter(COLUMNNAME == CRITERIA)
```

```{r 1.2response}

```

### 1.3

Write out the `ces_Yuba` object as a CSV file calling it "ces_Yuba.csv", using `write_csv()`:

```{r 1.3response}

```


# Practice on Your Own!

### P.1

Copy your code from problem 1.3 and modify it to write to the data directory inside your R Project. *Note: you may need to make a new folder named "data" if it doesn't already exist.*

```{r eval = FALSE, label = 'P.1response'}

```

### P.2
Write one of the objects in your Environment to your working directory in `rds` format. Call the file `my_variable.rds`.

```{r P.2response}

```

### P.3

Read the RDS file from your working directory back into your Environment. Call the file `z`.

```{r P.3response}

```
