---
title: "Data Output Lab - Key"
output: html_document
editor_options: 
  chunk_output_type: console
---

In this lab you can use the interactive console to explore or **Knit** the document. 
Remember anything you type here can be "sent" to the console with Cmd-Enter (OS-X) or Ctrl-Enter (Windows/Linux) in an R code chunk.


# Part 1

### 1.1

Read in the SARS-CoV-2 wastewater data from URL 
https://daseh.org/data/SARS-CoV-2_Wastewater_Data.csv
and assign it to an object named `covid`. 

```
# General format
library(readr)
# OBJECT <- read_csv(FILE)
```

```{r 1.1response}
library(tidyverse)
covid <- read_csv(file = "https://daseh.org/data/SARS-CoV-2_Wastewater_Data.csv")
```

### 1.2

Filter the dataset so that the "reporting_jurisdiction" column is equal to "Maryland". Store the modified dataset as `covid_filtered`.

```
# General format
NEW_OBJECT <- OBJECT %>% filter(COLUMNNAME == CRITERIA)
```

```{r 1.2response}
covid_filtered <- covid %>% filter(reporting_jurisdiction == "Maryland")
```

### 1.3

Write out the `covid_filtered` object as a CSV file calling it "covid_filtered.csv", using `write_csv()`:

```{r 1.3response}
write_csv(covid_filtered, file = "covid_filtered.csv")
```


# Practice on Your Own!

### P.1

Copy your code from problem 1.3 and modify it to write to the data directory inside your R Project. *Note: you may need to make a new folder named "data" if it doesn't already exist.*

```{r eval = FALSE, label = 'P.1response'}
getwd()
dir.create("data")
write_csv(covid_filtered, file = "data/covid_filtered.csv")
```

### P.2

Write one of the objects in your Environment to your working directory in `rds` format. Call the file `my_variable.rds`.

```{r P.2response}
y <- c(10, 20, 30, 40, 50, 60)
write_rds(y, file = "my_variable.rds")
```

### P.3

Read the RDS file from your working directory back into your Environment. Call the file `z`.

```{r P.3response}
z <- read_rds(file = "my_variable.rds")
```
