---
title: "Mapping"
output:
  ioslides_presentation:
    css: ../../docs/styles.css
    widescreen: yes
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(comment = "")

install.packages(c("maps", "sf", "tigris", "units", "usmapdata", "usmap", "tidycensus"), repos = "https://cloud.r-project.org", dependencies = "Depends")
```

## Making maps in R

Maps can be tricky in R! There are many packages to choose from.

- `usmap` is compatible with ggplot
- `maps` is "Base R"
- Some require API keys (e.g., `ggmap`, `tidycensus`)
- Some are interactive (e.g., `leaflet`)

## Data formats - boundary data

```{r message=FALSE}
library(tidyverse)
head(map_data("county"))
```

## Data formats - boundary data

- **long**: Longitude (x-coordinate)
- **lat**: Latitude (y-coordinate)
- **group**: Identifies unique polygons (each county may have multiple polygons if it contains islands or complex borders).
- **order**: Sequence in which points should be connected to form the boundary (polygons).
- **region**: State name (e.g., "alabama").
- **subregion**: County name within the state (e.g., "autauga").

## Data formats - sf data {.codesmall}

```{r message=FALSE}
library(usmapdata)
head(us_map("county"))
```

## Data formats - sf data {.codesmall}

This data is "Simple Feature" (sf) data used for spatial analysis. 

These objects store geometric shapes (like points, lines, or polygons) along with associated attributes (metadata).

the `geom` column is a MULTIPOLYGON — a geometry type representing complex shapes, which may consist of multiple polygons (e.g., islands or non-contiguous regions).

## `ggplot` has spatial functions{.codesmall}

`geom_polygon()` works with boundary data

```{r message=FALSE}
library(tidyverse) # `map_data()` from ggplot2
library(maps) # `us.cities` data

wa_county <- map_data("county") %>% filter(region == "washington")
wa_cities <- us.cities %>% filter(country.etc == "WA")

plot_1 <- 
  ggplot() + 
  geom_polygon(data = wa_county, aes(x = long, y = lat, group = group), color = "black", fill = NA) +
  geom_point(data = wa_cities, aes(x = long, y = lat)) +
  labs(title = "Washington State Cities", x = "longitude", y = "latitude") +
  coord_fixed(1.3)
```

## `ggplot` has spatial functions{.codesmall}

```{r echo=FALSE, message=FALSE}
plot_1
```

## `maps` package is more similar to Base R

```{r eval=FALSE}
library(maps) # `map` and `map.cities` functions and `us.cities` data

map('county', region = 'washington', col = "#5E610B")
map.cities(us.cities, country="WA", col = "#642EFE", cex = 2)
title(main = "Washington State Cities")
```

## `maps` package is more similar to Base R

```{r echo=FALSE, message=FALSE}
library(maps) # `map` and `map.cities` functions and `us.cities` data

map('county', region = 'washington', col = "#5E610B")
map.cities(us.cities, country="WA", col = "#642EFE", cex = 2)
title(main = "Washington State Cities")
```

## `usmap` is compatible with ggplot

```{r message=FALSE}
library(tidyverse)
library(usmap) # `countypop` data and the `plot_usmap()` function

wa_dat <- countypop %>% filter(abbr == "WA")

plot_3 <-
  plot_usmap(data = wa_dat, values = "pop_2022", include = c("WA")) +
  scale_fill_continuous() +
  theme(legend.position = "right")
```

## `usmap` is compatible with ggplot

```{r echo=FALSE, message=FALSE}
plot_3
```

## It can get complicated! `ggplot` fill by county {.codesmall}

Sometimes a lot of cleanup is needed to join boundary data with attributes of interest!

```{r message=FALSE}
library(tidyverse)
library(usmap) # `countypop` data
library(maps) # `us.cities` data

# Get county boundaries
wa_county <- map_data("county") %>% filter(region == "washington")

# Get county-level ("subregion") population
wa_dat <- countypop %>% filter(abbr == "WA") %>%
  mutate(subregion = tolower(str_remove(county, " County"))) %>%
  group_by(subregion) %>% summarize(pop_2022 = sum(pop_2022))

# Combine the data
wa_complete <- wa_county %>% inner_join(wa_dat)

# Get WA cities and their coordinates
wa_cities <- us.cities %>% filter(country.etc == "WA")
```

## It can get complicated! `ggplot` fill by county {.codesmall}

Sometimes a lot of cleanup is needed to join boundary data with attributes of interest!

```{r message=FALSE}
# Step 2: create the plot
plot_4 <-
  ggplot() + 
  geom_polygon(data = wa_complete, aes(x = long, y = lat, group = group, fill = pop_2022)) +
  geom_point(data = wa_cities, aes(x = long, y = lat), color = "red") +
  labs(title = "Washington State Population and Cities, 2022", x = "longitude", y = "latitude") +
  coord_fixed(1.3)
```

## It can get complicated! `ggplot` fill by county {.codesmall}

```{r echo=FALSE, message=FALSE}
plot_4
```

## `tidycensus` is helpful for tract level {.codesmall}

Use `geom_sf()` function with SF data.

```{r message=FALSE, warning=FALSE}
library(tidyverse) # `geom_sf()` from ggplot2
library(tidycensus) # `get_acs()` function for American Community Survey data

wa_income <- get_acs(
  geography = "tract", 
  variables = "B19013_001", # Income key 
  state = "WA", 
  year = 2022,
  geometry = TRUE
)
```

## `tidycensus` is helpful for tract level {.codesmall}

```{r message=FALSE}
ggplot(data = wa_income, aes(fill = estimate)) + 
  geom_sf()
```

## Tips for Mapping in R

1. Know the functions: make sure your data going into plotting functions is similar to 

2. Data Structure: Ensure column names match between datasets for join() operations

    - e.g., `subregion` needs to align in both `wa_county` and `wa_dat` to make `wa_complete`.
    - Make sure all datasets (like counties and cities) use the same geographic system, such as longitude-latitude pairs.

3. Clean Data to make life easier

    - Use functions like tolower() and str_remove() to standardize text (e.g., removing "County").
    - Group and summarize data when plotting aggregates, like population by county.

## More resources

https://rfortherestofus.com/2024/06/us-maps

https://ggplot2-book.org/maps

https://walker-data.com/tidycensus/articles/spatial-data.html

https://walker-data.com/census-r/mapping-census-data-with-r.html

https://jtr13.github.io/cc19/different-ways-of-plotting-u-s-map-in-r.html
