---
title: "Manipulating Data in R Lab - Key"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Some data in this lab comes from the OCS "Exploring CO2 emissions across time" activity (https://www.opencasestudies.org/ocs-bp-co2-emissions/).

Additional data about climate change disasters can be found at "https://daseh.org/data/Yearly_CC_Disasters_total_affected.csv". This dataset shows the total number of people affected by climate disasters by country and year.

```{r message=FALSE}
library(tidyverse)
```

# Part 1

### 1.1

Open the `yearly_co2_emissions` dataset. Use `read_csv()` from the `tidyverse` / `readr` package. You can download the data or use this URL directly: https://daseh.org/data/Yearly_CO2_Emissions_1000_tonnes.csv

Check out the data to understand the format.

```{r 1.1response}
co2 <- read_csv("https://daseh.org/data/Yearly_CO2_Emissions_1000_tonnes.csv")
head(co2)
```

### 1.2

Create a new dataset `co2_long` that does `pivot_longer()` on all columns except "country". Remember that `!country` means all columns except "country". 

Reassign to `co2_long`.

```
# General format
new_data <- old_data |> pivot_longer(cols = colname(s))
```

```{r 1.2response}
co2_long <- co2 |> pivot_longer(cols = !country)
```

### 1.3

Using `co2_long`, filter the "country" column so it only includes values from Indonesia and Canada. **Hint**: use `filter()` and `%in%`. 

Name the new object `co2_long_sub`.

```
# General format
new_data <- old_data |> filter(colname %in% c(...))
```

```{r 1.3response}
co2_long_sub <- co2_long |>
  filter(country %in% c("Indonesia", "Canada"))
```

### 1.4

Use `pivot_wider()` to reshape `co2_long_sub`. Use "country" for the `names_from` argument. Use "value" for the `values_from` argument. 

Call this new data `co2_wide`. 

```
# General format
new_data <- old_data |> pivot_wider(names_from = column1, values_from = column2)
```

```{r 1.4response}
co2_wide <- co2_long_sub |>
  pivot_wider(
    names_from = country,
    values_from = value
  )
```

### 1.5

Using `co2_wide`, drop all NA values using `drop_na()`.

Reassign to `co2_wide`. Compare the years - what conclusions can you draw, comparing the two countries?

```{r 1.5response}
co2_wide <- co2_wide |> drop_na()
tail(co2_wide)
# Indonesia emissions are lower to start, but become closer in value to Canada over time. However, Indonesia has many more people!
```


# Practice on Your Own!

### P.1

Take the code from Questions 1.1 - 1.5. Chain all of this code together using the pipe ` |> `. Call your data `co2_compare`.

```{r P.1response}
co2_compare <-
  read_csv("https://daseh.org/data/Yearly_CO2_Emissions_1000_tonnes.csv") |>
  pivot_longer(cols = !country) |>
  filter(country %in% c("Indonesia", "Canada")) |>
  pivot_wider(names_from = country, values_from = value) |> 
  drop_na()
  
co2_compare
```

### P.2

Modify the code from Question P.1:

-  Select different countries to compare
-  Call your data `co2_compare2`

```{r P.2response}
co2_compare2 <-
  read_csv("https://daseh.org/data/Yearly_CO2_Emissions_1000_tonnes.csv") |>
  pivot_longer(cols = !country) |>
  filter(country %in% c("Brazil", "Mexico")) |>
  pivot_wider(names_from = country, values_from = value) |> 
  drop_na()

co2_compare2
```

### P.3

Add on to Question P.1 to add a new column. The new column values for each line should be "TRUE" if Canada has greater emissions than Indonesia for that year.

```{r P.3response}
co2_compare <-
  co2_compare |> 
  mutate(Canada_greater = Canada > Indonesia)

co2_compare
```


# Part 2

### 2.1 

Using the `co2_long` object, rename the columns using `rename()`:

- change `name` to `year` 
- change `value` to `emissions`

Reassign to `co2_long`.

```
# General format
new_data <- old_data |> rename(newcol1 = oldcol1, newcol2 = oldcol2)
```

```{r 2.1response}
co2_long <- co2_long |> rename(year = name, emissions = value)
```

### 2.2 

Read in the climate change disasters dataset from the URL below. Save the dataset as an object called `cc`.

"https://daseh.org/data/Yearly_CC_Disasters_total_affected.csv"

This dataset shows the total number of people affected by climate disasters by country and year.

```{r 2.2response}
cc <- read_csv("https://daseh.org/data/Yearly_CC_disasters_total_affected.csv")
```

### 2.3

For the `co2_long` dataset, use `mutate()` to make the "year" column numeric.

Reassign to `co2_long`.

```{r 2.3response}
co2_long <- co2_long |> mutate(year = as.numeric(year))
```

### 2.4

Join the datasets `co2_long` and `cc` by "year" and "country". Keep all rows, even if there is no match. How many rows are there?

Call the dataset `fj`.

```
# General format
full_join(data1, data2, by = "")
```

```{r 2.4response}
fj <- full_join(co2_long, cc, by = c("year", "country"))
nrow(fj)
# 55,823 rows
```

### 2.5

Join the datasets `co2_long` and `cc` by "year" and "country". Keep only rows in both datasets. How many rows are there?

Call the dataset `ij`.

```
# General format
inner_join(data1, data2, by = "")
```

```{r 2.5response}
ij <- inner_join(co2_long, cc, by = c("year", "country"))
nrow(ij)
# 4,585 rows
```


# Practice on Your Own!

### P.4

Inspect `fj` and `ij`. Why is there such a big difference between the row counts?

```{r P.4response}
# There are many more years in the "co2_long" dataset than are present in the "cc" dataset. Could check out range to do this:
co2_long |> pull(year) |> range()
cc |> pull(year) |> range()

# Additionally, there are several indexes that don't match for country
# e.g., "Afghanistan" does not match with "Afghanistan, Islamic Rep. of"
# Perhaps the dataset could be cleaned up to account for these!
```

### P.5

What happens if you leave out step 2.3? Try running all chunks, but place a `#` on line 183. 

```{r P.5response}
# the year column in co2_long is a character type if we don't take the step in 2.3. This prevents joining as the columns joined must be the same class.
```

