---
title: "Manipulating Data in R Lab - Key"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Some data in this lab comes from the OCS "Exploring CO2 emissions across time" activity (https://www.opencasestudies.org/ocs-bp-co2-emissions/.

Additional data about climate change disasters can be found at "https://daseh.org/data/Yearly_CC_Disasters.csv".

```{r message=FALSE}
library(tidyverse)
```

# Part 1

### 1.1

Open the `yearly_co2_emissions` dataset. Use `read_csv()` from the `tidyverse` / `readr` package. You can download the data or use this URL directly: https://daseh.org/data/Yearly_CO2_Emissions_1000_tonnes.csv

Check out the data to understand the format.

```{r 1.1response}
co2 <- read_csv("https://daseh.org/data/Yearly_CO2_Emissions_1000_tonnes.csv")
head(co2)
```

### 1.2

Create a new dataset "co2_long" that does `pivot_longer()` on all columns except "country". Remember that `!country` means all columns except "country". 

Reassign to co2_long.

```
# General format
new_data <- old_data %>% pivot_longer(cols = colname(s))
```

```{r 1.2response}
co2_long <- co2 %>% pivot_longer(cols = !country)
```

### 1.3

Using `co2_long`, filter the "country" column so it only includes values from Indonesia and Canada. **Hint**: use `filter` and `%in%`. 

Reassign to co2_long.

```
# General format
new_data <- old_data %>% filter(colname %in% c(...))
```

```{r 1.3response}
co2_long <- co2_long %>%
  filter(country %in% c("Indonesia", "Canada"))
```

### 1.4

Use `pivot_wider` to reshape "co2_long". Use "county" for the `names_from` argument. Use "value" for the `values_from` argument. Call this new data `co2_wide`. 

```
# General format
new_data <- old_data %>% pivot_wider(names_from = column1, values_from = column2)
```

```{r 1.4response}
co2_wide <- co2_long %>%
  pivot_wider(
    names_from = country,
    values_from = value
  )
```

### 1.5

Using `co2_wide`, drop all NA values using `drop_na()`.

Reassign to co2_wide. Compare the years - what conclusions can you draw?

```{r 1.5response}
co2_wide <- co2_wide %>% drop_na()
```

Tip: you can adjust scientific notation with `options(scipen=<n>)` 


# Practice on Your Own!

### P.1

Take the code from Questions 1.1 - 1.5. Chain all of this code together using the pipe ` %>% `. Call your data `co2_compare`.

```{r P.1response}
co2_compare <-
  read_csv("https://daseh.org/data/Yearly_CO2_Emissions_1000_tonnes.csv") %>%
  pivot_longer(cols = !country) %>%
  filter(country %in% c("Indonesia", "Canada")) %>%
  pivot_wider(names_from = country, values_from = value) %>% 
  drop_na()
  
co2_compare
```

### P.2

Modify the code from Question P.1:

-  Select different countries to compare
-  Call your data `co2_compare2`

```{r P.2response}
co2_compare2 <-
  read_csv("https://daseh.org/data/Yearly_CO2_Emissions_1000_tonnes.csv") %>%
  pivot_longer(cols = !country) %>%
  filter(country %in% c("Brazil", "Mexico")) %>%
  pivot_wider(names_from = country, values_from = value) %>% 
  drop_na()

co2_compare2
```

### P.3

Add on to Question P.1 to add a column. The column values for each line should be "TRUE" if Canada has greater emissions than Indonesia for that year.

```{r P.3response}
co2_compare <-
  co2_compare %>% 
  mutate(Canada_greater = Canada > Indonesia)

co2_compare
```


# Part 2

### 2.1

Open the `Yearly_CC_Disasters` dataset using the url below. Save the dataset as an object called "cc". rename the column "Country" to "country".

"https://daseh.org/data/Yearly_CC_Disasters.csv"


```{r 2.1response}
cc <- read_csv("https://daseh.org/data/Yearly_CC_Disasters.csv") %>%
  rename(country = Country)
```


### 2.2

Join the datasets "co2" and "cc" by "country". Keep all rows, even if there is no match. How many rows are there?

```
# General format
full_join(data1, data2, by = "")
```

```{r 2.2response}
full_join(co2, cc, by = "country")
# 1,032 rows
```


### 2.3

Join the datasets "co2" and "cc" by "country". Keep only rows that match both datasets. How many rows are there?

```
# General format
inner_join(data1, data2, by = "")
```

```{r 2.3response}
inner_join(co2, cc, by = "country")
# 628 rows
```


### 2.4

What countries are present in "co2" that are not present in "cc"? Use `anti_join`. What countries are present in "cc" that are not present in "co2"?

```
# General format
anti_join(data1, data2, by = "") %>% select(columnname)
```

```{r 2.4response}
anti_join(co2, cc, by = "country") %>% select(country)
anti_join(cc, co2, by = "country") %>% select(country) %>% distinct()
```


# Practice on Your Own!

### P.3

Take the code from 2.2 and save the output as an object "co2_cc". Filter the dataset. Filter so that you only keep Indonesia and Canada.

```{r P.3response}
co2_cc <- full_join(co2, cc, by = "country") %>%
  filter(country %in% c("Indonesia", "Canada"))
```


### P.4

Select:

* the "country" column
* data from the years 2014 originally in BOTH DATASETS (columns "CO2_2014" and `2014`)
* the "Indicator" column

Rename:

* emissions = CO2_2014
* disasters = `2014`

Reassign to "co2_cc".

```{r P.4response}
co2_cc <- co2_cc %>% select(country, CO2_2014, Indicator, `2014`) %>%
  rename(emissions = CO2_2014, disasters = `2014`)
```


### P.5

Use `stringr` to trim the piece of text, "Climate related disasters frequency, Number of Disasters: ", from the "Indicator" column. You will use the function `str_remove()` to do this. It works similarly to other `stringr` functions. Try to intuit how it works by using the documentation page (`?str_remove`).

```{r P.5response}
library(stringr)
co2_cc <- co2_cc %>% mutate(Indicator = str_remove(
  Indicator,
  "Climate related disasters frequency, Number of Disasters: "
))
```


### P.6

Pivot the dataset so that there are columns for country, emissions, and a column for each "Indicator".

```{r P.6response}
co2_cc %>% pivot_wider(
  names_from = Indicator,
  values_from = disasters
)
```
