---
title: "Data Input Lab"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Part 1

## Helpful tips before we start

### TROUBLESHOOTING: Common new user mistakes we have seen

- Check the file path -- is the file there?  
- Typos (R is **case sensitive**, `x` and `X` are different)  
- Open ended quotes, parentheses, and brackets     
- Deleting part of the code chunk  
- For any function, you can write `?FUNCTION_NAME`, or `help("FUNCTION_NAME")` to look at the help file

### 1.1

Set up your R Project.

> File, New Project or click the new project button
> New Directory
> New Project
> Type a name and choose a location
> Check that the folder is there!

Check out our resource here: https://daseh.org/resources/R_Projects.html

### 1.2 

Load the package by adding "library(tidyverse)" below and running the code.

```{r 1.2response}

```

### 1.3

Use the manual import method (File > Import Dataset > From Text (`readr`)) to read in Haloacetic Acids (HAA5) Exposure for WA Populations on Public Water Systems data from this URL:

https://daseh.org/data/HAA5_Exposure_for_WA_Public_Water_Systems_data.csv.

These data were collected by the Washington Tracking Network. You can learn more about the data here: https://fortress.wa.gov/doh/wtn/WTNPortal/#!q0=674

### 1.4

What is the dataset object called? You can find this information in the Console or the Environment. Enter your answer as a comment using `#`.

```{r 1.4response}

```

### 1.5

Preview the data by examining the Environment. How many observations and variables are there? Enter your answer as a comment using `#`.

```{r 1.5response}

```

# Practice on Your Own!

### P.1

Download the data and move the file to your project folder. Import the data by browsing for the file on your computer.

> *Download the data*
> *Put data in the project folder*
> File, Import Dataset, From Text (`readr`)
> browse for the file
> click "Update" and "Import"


# Part 2

### 2.1

Read Haloacetic Acids (HAA5) Exposure for WA Populations on Public Water Systems data from https://daseh.org/data/HAA5_Exposure_for_WA_Public_Water_Systems_data.csv.
and assign it to an object named `exposure`. Use the code structure below.

```
# General format
OBJECT <- read_csv(FILE)
```

```{r 2.1response}

```

### 2.2

Take a look at the data. Do these data objects (`HAA5_Exposure_for_WA_Public_Water_Systems_data` and `exposure`) appear to be the same? Why or why not?

```{r 2.2response}

```

### 2.3

Learn your working directory by running `getwd()`. This is where R will look for files unless you tell it otherwise.

```{r 2.3response}

```


# Practice on Your Own!

### P.2

Run the following code - is there a problem? How do you know?

```{r}
exposure <- read_delim("https://daseh.org/data/HAA5_Exposure_for_WA_Public_Water_Systems_data.csv", delim = "\t")
exposure
```

```{r P.2response}

```

### P.3

Try reading in some data on your computer using any method we discussed!
