---
title: 'Introduction to R: Homework 3'
output:
  html_document: default
---

### **Instructions**



Homework is optional, but we recommend it so you can get the most out of this course.

```{r initiatePackages, message=FALSE}
## you can add more, or change...these are suggestions
library(tidyverse)
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
```


### **Problem Set**

1\. Bring the following dataset into R. 

-  The dataset is located at: https://daseh.org/data/Baltimore_surface_water_quality_data.csv. 
-  You can use the link, download it, or use whatever method you like for getting the file. 
-  Once you get the file, read the dataset in using `read_csv()` and assign it the name "water".

```{r 1response}

```


2\. Run the `colnames()` function to take a look at the dataset column names. Rename the third column of "water" to "GPS_North" using the `rename()` function in `dplyr`. Then rename the fourth column to "GPS_West".  

```{r 2response}

```


3\. Convert the "datetime" column to POSIXct format using the `mdy_hm()` functions from the `lubridate` package. Then create the variables "Month" and "Year" by extracting the month and year from the `datetime` column. (HINT: use mutate and `year(datetime_fixed)` to create your new variables.)


```{r 3response}

```

4\. Separate `datetime_fixed` into two separate columns: "date" and "time" using the `separate()` function. Replace the original "water" object by calling the new dataset "water".

```{r 4response}

```

5\. What is the class for "Result"? Can you take the mean value? Why or why not? What can you do to change the format of "Result" to numeric? 

```{r 5response}

```

6\. Replace the below detectable level values in "Results" with a measure that you choose. You should use a combination of `str_replace`,  `pull()` (because stringr package functions work on vectors not dataframes!) and `mutate()`. Replace the original "water" object by calling the new dataset "water".

-  The `pattern =` to find the starting parenthesis is "^<"
-  The `replacement =` depends on how you choose to replace the values

```{r 6response}

```

7\. Use `as.numeric()` to convert "Result" to numeric class. Then convert "Parameter" to a factor with `as.factor`.

```{r 7response}

```

8\. How many different measurements (levels) are stored in the Parameter column?

```{r 8response}

```

9\. Use the `pct_complete()` function in the `naniar` package to determine the percent missing data in "Results". You might need to load and install `naniar`!

```{r 9response}

```

10\. Are there any parameter levels that have a complete record in "water" across all years? **Hint:** Create a new dataframe using `group_by` and `summarize` to count the number of NAs in "Result" for each level in "Parameter", then filter out rows with more than 0.

```{r 10response}

```

11\. Subset "water"  so that it only contains results for the years 2010 - 2024, using `&` and the filter function. Make sure to include both the years 2010 and 2024. Confirm your filtering worked by looking at the range of "Year". Assign this subsetted tibble to "water_subset".

```{r 11response}

```


12\. Subset "water_subset"  so that it only contains results from the tests for dissolved oxygen, pH, and water temperature, using %in% and the filter function. Make sure to reassign to "water_subset".

```{r 12response}

```


13\. Load the new dataset "Baltimore_rainfall_HamiltonAve.csv" into R.  

-  The dataset is located at: https://daseh.org/data/Baltimore_rainfall_HamiltonAve.csv. 
-  You can use the link, download it, or use whatever method you like for getting the file. 
-  Once you get the file, read the dataset in using `read_csv()` and assign it the name "rainfall".

This dataset contains measured precipitation by month, as collected near the Hamilton Ave testing location in Baltimore.

```{r 13response}

```

14\. We want only the rainfall measurements for the years 2010 - 2024 from this new dataset, as well as the month information. Subset the dataset using the `select()` function. Reassign this to `rainfall`. HINT: Make sure you remember to keep the column that contains information about the month!

```{r 14response}

```


15\. Reshape `rainfall` into long format using `pivot.longer`. The reshaped dataset should contain three columns: year ("Year"), month ("Month"), and amount of rainfall ("Precip"). ou should pivot all columns except "Month". **Hint:** listing `!COLUMN` or `-COLUMN` means everything except COLUMN. Assign the reshaped dataset to `rainfall_long`.

```{r 15response}

```

16\. We would like to join the rainfall measurements dataset with the Baltimore surface water quality dataset, but we need to do a bit more wrangling first. Because the rainfall measures were collected near the Hamilton Ave water testing site, let's pull only the Hamilton Ave data from the `water_subset` data, using `filter`. Assign this to `water_Ham`.

```{r 16response}

```

17\. Right-join "water_Ham" and "rainfall_long" by "Month" and "Year" and assign the joined dataset to the name "water_rain". Did this join attempt work? Why or why not?

```{r 17response, eval=FALSE}
water_rain <- water_Ham |> right_join(rainfall_long, by = c("Month", "Year"))

#This attempted join did not work, because the class of Month is not the same between the two datasets!
```

18\. We have formatted "Month" differently in each dataset. In "rainfall_long", "Month" is a character and has a three-letter abbreviation for each month. In "water_Ham", "Month" is designated by a number. Reformat "Month" in "rainfall_long" so that it matches the format in "water_Ham".

```{r 18response}

```


19\. NOW try to right-join "water_Ham" and "rainfall_long" by "Month" and "Year" and assign the joined dataset to the name "water_rain". Did this join attempt work?

```{r 19response, eval=FALSE}
water_rain <- water_Ham |> right_join(rainfall_long, by = c("Month", "Year"))

#This attempted join did not work either, because "Year" is a different format between both datasets too!
```

20\. Fix the differences between the classes of "Year" and right-join the two datasets to make "water_rain".

```{r 20response}

```


21\. Plot the relationship between pH and rainfall, colored by Month and faceted by Year. You can use esquisse or ggplot (whichever you prefer).

```{r 21response}

```
