---
title: 'Introduction to R: Homework 3 - Key'
output:
  html_document: default
---

### **Instructions**



Homework is optional, but we recommend it so you can get the most out of this course.

```{r initiatePackages, message=FALSE}
## you can add more, or change...these are suggestions
library(tidyverse)
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
```


### **Problem Set**

1\. Bring the following dataset into R. 

-  The dataset is located at: https://daseh.org/data/Baltimore_surface_water_quality_data.csv. 
-  You can use the link, download it, or use whatever method you like for getting the file. 
-  Once you get the file, read the dataset in using `read_csv()` and assign it the name "water".

```{r 1response}
water <- read_csv("https://daseh.org/data/Baltimore_surface_water_quality_data.csv")
```


2\. Run the `colnames()` function to take a look at the dataset column names. Rename the third column of "water" to "GPS_North" using the `rename()` function in `dplyr`. Then rename the fourth column to "GPS_West".  

```{r 2response}
colnames(water)
water <- water |> rename(GPS_North = `GPS Coordinate North`)
water <- water |> rename(GPS_West = `GPS Coordinate West`)
```


3\. Convert the "datetime" column to POSIXct format using the `mdy_hm()` functions from the `lubridate` package. Then create the variable "year" by extracting the year from the `datetime` column. (HINT: use mutate and `year(datetime_fixed)` to create your new variable.)


```{r 3response}
#converting the date format to POSIXct
water <- water |> mutate(datetime_fixed = mdy_hm(datetime))

#creating a new variable
water <- water |> mutate(year = year(datetime_fixed))

```

18\. Separate `datetime_fixed` into two separate columns: "date" and "time" using the `separate()` function. Replace the original "water" object by calling the new dataset "water".

```{r 18response}
water <- water |> separate(`datetime_fixed`, into = c("date", "time"), sep = " ")
```

4\. What is the class for "Result"? Can you take the mean value? Why or why not? What can you do to change the format of "Result" to numeric? 

```{r 4response}
water |> pull(Result) |> class()

# "Result" is of type "character". For some tests, the result is listed as below the detectable level, using a ">". R can only interpret this symbol as a character. In order to change "Result" to numeric, we need to change the values of the results that are below detectable levels. Some options are to change these to 0 or to NA. You can also change them on a case-by-case basis to a number just slightly smaller than the detectable level (for example, if the level is 3 ug, anything below this could be coded as 2.9 ug). Use your best judgment on how to recode these values.

```

19\. Replace the below detectable level values in "Results" with a measure that you choose. You should use a combination of `str_replace`,  `pull()` (because stringr package functions work on vectors not dataframes!) and `mutate()`. Replace the original "water" object by calling the new dataset "water".

-  The `pattern =` to find the starting parenthesis is "^<"
-  The `replacement =` depends on how you choose to replace the values

```{r 19response}
#replacing values with NA
water <- water |> mutate(Result = str_replace(pull(water, Result), pattern = "^<", replacement = NA_character_))

#replacing values with 0
#water <- water |> mutate(Result = str_replace(pull(water, Result), pattern = "^<", replacement = 0))
```

21\. Use `as.numeric()` to convert "Result" to numeric class. Then convert "Parameter" to a factor with `as.factor`.

```{r 21response}
water |>
  mutate(Result = as.numeric(Result)) 
water |> 
  mutate(Parameter = as.factor(Parameter))
```

xx\. How many different measurements (levels) are stored in the Parameter column?

```{r xxresponse}
water |> pull(Parameter) |> levels()

#There are 49 different types of measurements in this dataset.
```

5\. Use the `pct_complete()` function in the `naniar` package to determine the percent missing data in "Results". You might need to load and install `naniar`!

```{r 5response}
library(naniar)
pct_complete(water)
# "water" is 99.99913 percent complete if you chose to replace results below the detectable limit with 0. The dataset is 99.41319 if you instead chose to replace results below the detectable limit with NA.
```

6\. Are there any parameter levels that have a complete record in "water" across all years? **Hint:** Create a new dataframe using `group_by` and `summarize` to count the number of NAs in "Result" for each level in "Parameter", then filter out rows with more than 0.

```{r 6response}
na_counts <- water %>%
  group_by(Parameter) %>%
  summarise(na_count = sum(is.na(Result)))

na_counts |> filter(na_count == 0)

#If you replaced the values below detectable levels with NA, there are seven parameter levels (tests) that have no missing data. If you replaced the values below detectable levels with 0, all the parameter levels (tests) have complete data.
```


7\. Reshape the "complete" data to long form. 

- There should be a column for country ("country"), a column for year ("year"), and a column for the mortality value ("mortality"). 
- Use `pivot_longer()`. 
- You should pivot all columns except "country". 
- **Hint:** listing `!COLUMN` or `-COLUMN` means everything except COLUMN. 
- Assign the reshaped data to "long".

```{r 7response}
long <-
  pivot_longer(mort, -country, names_to = "year", values_to = "mortality")
# OR
long <-
  pivot_longer(mort, !country, names_to = "year", values_to = "mortality")
# OR
long <-
  pivot_longer(mort, starts_with(c("1", "2")), names_to = "year", values_to = "mortality")
```


8\. Bring an additional dataset into R. 

-  The dataset is tab-delimited and located at: https://daseh.org/data/country_pop.txt. 
-  You can use the link, download it, or use whatever method you like for getting the file. 
-  Once you get the file, read the dataset in using `read_tsv()` and assign it the name "pop".

```{r 8response}
pop <- read_tsv("https://daseh.org/data/country_pop.txt")
```


9\. Rename the second column in "pop" to "country" and the column "% of world population", to "percent". Use the `rename()` function. Don't forget to reassign the renamed data to "pop".

```{r 9response}
pop <- pop |>
  rename(
    country = `Country (or dependent territory)`,
    percent = `% of world population`
  )
```


10\. Sort the data in "pop" by "Population" from largest to smallest using `arrange()` and `desc()`. After sorting, `select()` "country" to create an one-column tibble of countries ordered by population. Assign this data the name "country_ordered".

```{r 10response}
country_ordered <- pop |>
  arrange(desc(Population)) |>
  select(country)
```


11\. Subset "long"  based on years 2000-2010, including 2000 and 2010 and call this "long_sub" using `&` or the `between()` function. Confirm your filtering worked by looking at the range of "year". If you're getting a strange error, make sure you created the "year" column in problem #7.

```{r 11response}
long_sub <- long |> filter(year >= 2000 & year <= 2010)
long_sub |>
  pull(year) |>
  range() # confirm it worked
```


12\. Further subset `long_sub`. You will filter for specific countries using `filter()` and the `%in%` operator. Only include countries in this list: `c("Venezuela", "Bahrain", "Estonia", "Iran", "Thailand", "Canada")`. Make sure to reassign to "long_sub".

```{r 12response}
long_sub <- long_sub |>
  filter(country %in% c("Venezuela", "Bahrain", "Estonia", "Iran", "Thailand", "Canada"))
```


13\. Use `pivot_wider()` to turn the "year" column of "long_sub" into multiple columns, each representing a different year. Fill values (`values_from=`) with "mortality". Assign this pivoted dataset the name "mort_sub".

```{r 13response}
mort_sub <- long_sub |>
  pivot_wider(id_cols = country, names_from = year, values_from = mortality)
```


14\. Using "country_ordered" and "mort_sub", `right_join()` the two datasets by "country". Use the pipe `|>` to join this dataset to "pop", keeping only the data on the lefthand side of the join. Call this "joined".

```{r 14response}
joined <- country_ordered |>
  right_join(mort_sub, by = "country") |>
  left_join(pop, by = "country")
```


15\. The values in the table are percentages of the total population (not proportion). 

-  Create a new column called "mort_count" that estimates the total number of child deaths per year based on the total population. You can use (a) any year or (b) an average of all years to make your calculation. Whichever you choose, justify your choice. 
-  Finally, `select()` only "country", "Population", and "mort_count" and view the data.

> Justification is just for fun. The main point is that decisions in your analysis should depend on your reasoning not how many lines of code it takes :)

```{r 15response}
# Use 2010: There appears to be a downward trend in mortality rates, so using 2010 could be the most accurate for future years.
joined |>
  mutate(mort_count = Population * `2010` / 100) |>
  select(country, Population, mort_count)

# OR

# Use an average 2000-2010: Using the average is more robust to fluctuations and uncertainty in the coming years.
avg_mort <- rowMeans(joined |> select(starts_with("2")))
joined <- joined |> mutate(avg_pct_2000 = avg_mort)
joined |>
  mutate(mort_count = Population * avg_pct_2000 / 100) |>
  select(country, Population, mort_count)
```



