---
title: "Data Summarization Lab"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1

We'll again use the CalEnviroScreen dataset for the lab. Load the `tidyverse` package and the dataset, which can be found at 	https://daseh.org/data/CalEnviroScreen_data.csv. Name the dataset `ces`.


```{r, echo = TRUE, message=FALSE, error = FALSE}
library(tidyverse)

ces <- read_csv(file = "https://daseh.org/data/CalEnviroScreen_data.csv")
```

### 1.1 

How many observations/rows are in the `ces` data set? You can use `dim()` or `nrow()` or examine the Environment.

```{r 1.1response}

```

### 1.2

The `TotalPop` column includes information about the population for each census tract as of the 2010 census.

NOTE: A census tract a small, relatively permanent area within a county used to present data from the census. Each row in the `ces` dataset corresponds to a single census tract. See https://www2.census.gov/geo/pdfs/education/CensusTracts.pdf

What was the total population in the dataset based on the 2010 census? (use `sum()` and the `TotalPop` column)

```{r 1.2response}

```

### 1.3

What was the largest population, according to the 2010 census? Use `summarize` and `max`.

```
# General format 
DATA_TIBBLE |> 
    summarize(SUMMARY_COLUMN_NAME = FUNCTION(SOURCE_COLUMN))
```

```{r 1.3response}

```

### 1.4

Modify your code from 1.3 to add the smallest population among census tracts. Use `min` in your `summarize` function. 

*HINT:* Make sure you make your `SUMMARY_COLUMN_NAME` different for the max and the min summaries! What do you think would happen if you tried to make `SUMMARY_COLUMN_NAME` the same name for both statistics?

```
# General format 
DATA_TIBBLE |> 
    summarize(SUMMARY_COLUMN_NAME = FUNCTION(SOURCE_COLUMN),
              SUMMARY_COLUMN_NAME = FUNCTION(SOURCE_COLUMN)
    )
```

```{r 1.4response}

```


# Practice on Your Own!

### P.1

Summarize the `ces` data to get the mean of both the `TotalPop` and `Pesticides` columns. Make sure to remove `NA`s.

`Pesticides`: Total pounds of selected active pesticide ingredients  used in production-agriculture per square mile. The higher the number, the greater the amount of pesticides have been used on agricultural sites

```
# General format 
DATA_TIBBLE |> 
    summarize(SUMMARY_COLUMN_NAME = OPERATOR(SOURCE_COLUMN, na.rm = TRUE),
              SUMMARY_COLUMN_NAME = OPERATOR(SOURCE_COLUMN, na.rm = TRUE)
    )
```

```{r P.1response}

```

### P.2

Given that parts of California are heavily agricultural, and the max value for the `Pesticides` variable is 80811, why might the average value be so low??

```{r P.2response}

```

### P.3

Filter any zeros from the `Pesticides` column out of `ces`. Use `filter()`. Assign this "cleaned" dataset object the name `ces_pest`.

```
# General format 
DATA_TIBBLE |> filter(LOGICAL_COMPARISON)
```

```{r P.3aresponse}

```

How many census tracts have pesticide values greater than 0?

```{r P.3bresponse}

```


# Part 2

### 2.1

How many census tracts are present in each California county? Use `count()` on the column named `CaliforniaCounty`. Use `ces` as your input data.

```{r 2.1response}

```

### 2.2

Let's break down the count further. Modify your code from question 2.1 to count census tracts by County AND ZIP code. Use `count()` on the columns named `CaliforniaCounty` and `ZIP`.

```{r 2.2response}

```

This isn't the only way we can create this table in R. Let's look at another way to build it.

### 2.3

How many census tracts are there in each county? Use `group_by()`, `summarize()`, and `n()` on the column named `CaliforniaCounty`.

```
# General format 
DATA_TIBBLE |> 
    group_by(GROUPING_COLUMN_NAME) |> 
    summarize(SUMMARY_COLUMN_NAME = n())
```

```{r 2.3response}

```

### 2.4

Modify your code from 2.3 to also group by `ZIP`.

```{r 2.4response}

```


# Practice on Your Own!

### P.4

Modify code from 2.3 (the code that only groups by county) to also summarize by total population (`TotalPop`) per group. In your summarized output, make sure you call the new summarized average total population variable (column name) "mean". In other words, the head of your output should look like:

```
# A tibble: 58 × 3
   CaliforniaCounty count  mean
   <chr>            <int> <dbl>
 1 "Alameda "         360 4602.
...
```
(In the above table, remember that the "count" column is counting the number of census tracts.)

```{r P.4response}

```

### P.5

Take your code from the above question P.4 and do the following:

- Add another pipe (`|>`)
- Add the `arrange()` to sort the output by the summarized column "mean".

```{r P.5response}

```
