---
title: "Functions Lab - Key"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1

Load the `tidyverse` package.

```{r message=FALSE}
library(tidyverse)
```

### 1.1

Create a function that: 

* Takes one argument, a vector.
* Returns the sum of the vector and then squares the result. 
* Call it "sum_squared". 
* Test your function on the vector `c(2,7,21,30,90)` - you should get the answer 22500.
* Format is `NEW_FUNCTION <- function(x, y) x + y`

```{r 1.1response}
nums <- c(2, 7, 21, 30, 90)

sum_squared <- function(x) sum(x)^2
sum_squared(x = nums)

sum_squared <- function(x) {
  out <- sum(x)^2
  return(out)
}
sum_squared(x = nums)
```

### 1.2

Create a function that:

* takes two arguments, (1) a vector and (2) a numeric value. 
* This function tests whether the number (2) is contained within the vector (1). **Hint**: use `%in%`. 
* Call it `has_n`. 
* Test your function on the vector `c(2,7,21,30,90)` and number `21` - you should get the answer TRUE.

```{r 1.2response}
nums <- c(2, 7, 21, 30, 90)
a_num <- 21

has_n <- function(x, n) n %in% x
has_n(x = nums, n = a_num)
```

### 1.3

Amend the function `has_n` from question 1.2 so that it takes a default value of `21` for the numeric argument.

```{r 1.3response}
nums <- c(2, 7, 21, 30, 90)
a_num <- 21

has_n <- function(x, n = 21) n %in% x
has_n(x = nums)
```

### P.1

Create a function for the CalEnviroScreen Data. 

* Read in (https://daseh.org/data/CalEnviroScreen_data.csv)
* The function takes an argument for a column name. (use `{{col_name}}`)
* The function creates a ggplot with `{{col_name}}` on the x-axis and `Poverty` on the y-axis.
* Use `geom_point()`
* Test the function using the `Lead` column and `HousingBurden` columns, or other columns of your choice.

```{r P.1response}
ces <- read_csv("https://daseh.org/data/CalEnviroScreen_data.csv")

plot_ces <- function(col_name){
  ggplot(data = ces, aes(x = {{col_name}}, y = Poverty)) +
    geom_point()
}

plot_ces(Lead)
plot_ces(HousingBurden)
```


# Part 2

### 2.1

Read in the CalEnviroScreen from https://daseh.org/data/CalEnviroScreen_data.csv. Assign the data the name "ces".

```{r message = FALSE, label = '2.1response'}
ces <- read_csv("https://daseh.org/data/CalEnviroScreen_data.csv")
# If downloaded
# ces <- read_csv("CalEnviroScreen_data.csv")
```

### 2.2

We want to get some summary statistics on water contamination. 

* Use `across` inside `summarize`.
* Choose columns about "water". **Hint**: use `contains("water")` inside `across`. 
* Use `mean` as the function inside of `across`.
* Remember that `NA` values can influence calculations.

```
# General format
data %>%
  summarize(across(
    {vector or tidyselect},
    {some function}
  ))
```

```{r 2.2response}
ces %>%
  summarize(across(
    contains("water"),
    mean
  ))

# Accounting for NA
ces %>%
  summarize(across(
    contains("water"),
    function(x) mean(x, na.rm = T)
  ))
```

### 2.3

Convert all columns that are percentiles into proportions.

* Use `across` and `mutate`
* Choose columns that contain "Pctl" in the name. **Hint**: use `contains("Pctl")` inside `across`.
* Use an anonymous function ("function on the fly") to divide by 100 (`function(x) x / 100`). 
* Check your work - It will also be easier if you `select(contains("Pctl"))`.

```
# General format
data %>%
  mutate(across(
    {vector or tidyselect},
    {some function}
  ))
```

```{r 2.3response}
ces %>%
  mutate(across(
    contains("Pctl"),
    function(x) x / 100
  )) %>%
  select(contains("Pctl"))
```


# Practice on Your Own!

### P.2

Use `across` and `mutate` to convert all columns starting with the string "PM" into a binary variable: TRUE if the value is greater than 10 and FALSE if less than or equal to 10. 

* **Hint**: use `starts_with()` to select the columns that start with "PM". 
* Use an anonymous function ("function on the fly") to do a logical test if the value is greater than 10.
* A logical test with `mutate` (x > 10) will automatically fill a column with TRUE/FALSE.

```{r P.2response}
ces %>%
  mutate(across(
    starts_with("PM"),
    function(x) x > 10
  )) %>% 
  glimpse() # add glimpse to view the changes
```

### P.3

Take your code from previous question and assign it to the variable `ces_dat`.

- Create a ggplot where the x-axis is `Asthma` and the y-axis is `PM2.5`.
- Add a boxplot (`geom_boxplot()`)
- Change the `labs()` layer so that the x-axis is "ER Visits for Asthma: PM2.5 greater than 10"

```{r P.3response}
ces_dat <-
  ces %>%
  mutate(across(
    starts_with("PM"),
    function(x) x > 10
  ))

ggplot(data = ces_dat, aes(x = `Asthma`, y = `PM2.5`)) +
  geom_boxplot() +
  labs(x = "ER Visits for Asthma: PM2.5 greater than 10")

# Make everything a function if you like!
ces_boxplot <- function() {
  ces %>%
    mutate(across(
      starts_with("PM"), 
      function(x) x > 10
    )) %>% 
    ggplot(aes(x = `Asthma`, y = `PM2.5`)) +
    geom_boxplot() +
    labs(x = "ER Visits for Asthma: PM2.5 greater than 10")
}

ces_boxplot()
```
