---
title: "wasteWater_lab"
output: html_document
date: "2025-09-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Wastewater data set.
We will continue to use the wastewater data set to practice our R skills.
You can download the data here:
https://www.cdc.gov/wcms/vizdata/NCEZID_DIDRI/SC2/nwsssc2stateactivitylevelDL.csv



# Explore the data set:
Remember that  the function `head(n)` will print the first n rows of your dataframe, and the function `summary()` will give you 
snapshots of each numeric column while skipping character columns:
```{r}
head(dat)
summary(dat)
```

## Use your R knowledge to answer to the following questions:
- Are all states represented here? (hint: use the function `unique` to see unique elements of each column)
- What is the maximum wastewater value observed in a single state in this dataset (the wastewater value is given in the column named `State/Territory_WVAL`? Remember to include the parameter na.rm=TRUE to the max function!
- What is the minimum  wastewater value observed in a single state in this dataset?
- What is the maximum wastewater value observed at the national level (this is given in the column named `National_WVAL`)?
- Can you explain why the maximum value observed at the state level is much higher than the one at the national level?

## Answers:
```{r}
unique(dat$`State/Territory`)
dat |> pull(`State/Territory_WVAL`) |> max(na.rm = TRUE)
dat |> pull(`State/Territory_WVAL`) |> min(na.rm = TRUE)
dat |> pull(`National_WVAL`) |> max(na.rm = TRUE)

```
The national average is lower than the state average because it is the average across all states. 