---
title: "Wastewater_lab_key"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Part 1
Start at the beginning: load the packages.

```{r, message=FALSE}
library(tidyverse)
library(ggplot2)
```



We will continue to use the wastewater data set to practice our data science skills.
You can download the data here:
http://daseh.org/data/wastewater_COVID-19_State_and_Territory_Trends.csv.

<!--```{r, message = FALSE}
dat <- read_csv("https://daseh.org/data/SARS-CoV-2_Wastewater_Data.csv")
```
--> 
```{r, message = FALSE}
dat <- read_csv("https://www.cdc.gov/wcms/vizdata/NCEZID_DIDRI/SC2/nwsssc2stateactivitylevelDL.csv")
head(dat)
```

# Explore the data set:
Remember that  the function `head(n)` will print the first n rows of your dataframe, and the function `summary()` will give you 
snapshots of each numeric column while skipping character columns:
```{r}
head(dat)
summary(dat)
```

### 1.1

Use your R knowledge to answer to the following questions:

1. Are all states represented here? (hint: use the function `unique` to see unique elements of each column)
2. What is the mean wastewater value observed in a single state in this dataset (the wastewater value is given in the column named `State/Territory_WVAL`? Remember to include the parameter `na.rm=TRUE` to the mean function!
3. What is the minimum  wastewater value observed in a single state in this dataset?
4. What are the maximum wastewater values observed at the national level (this is given in the column named `National_WVAL`) and state level?
5. Can you explain why the maximum value observed at the state level is much higher than the one at the national level?

```{r 1.1response}
unique(dat$`State/Territory`)
dat |> pull(`State/Territory_WVAL`) |> mean(na.rm = TRUE)
dat |> pull(`State/Territory_WVAL`) |> min(na.rm = TRUE)
dat |> pull(`State/Territory_WVAL`) |> max(na.rm = TRUE)
dat |> pull(`National_WVAL`) |> max(na.rm = TRUE)

#The national maximum is lower than the state maximum because it is the average across all states. 
```

