---
title: "Data Class Lab - Key"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Part 1

### 1.1

Load the package we will use in this lab. 

```{r}
library(tidyverse)
```

Create some data to work with by running the following code chunk.

```{r}
set.seed(1234)

int_vect <- rep(seq(from = 1, to = 10), times = 3)
rand_vect <- sample(x = 1:30, size = 30, replace = TRUE)
TF_vect <- rep(c(TRUE, TRUE, FALSE), times = 10)
TF_vect2 <- rep(c("TRUE", "TRUE", "FALSE"), times = 10)
```

### 1.2

Determine the class of each of these new objects.

```{r 1.2response}
class(int_vect) # [1] "integer"
class(rand_vect) # [1] "integer"
class(TF_vect) # [1] "logical"
class(TF_vect2) # [1] "character"
```

### 1.3

Are `TF_vect` and `TF_vect2` different classes? Why or why not?

```{r 1.3response}
# Yes!
# Logical vectors do not have quotes around `TRUE` and `FALSE` values.
```

### 1.4

Create a tibble combining these vectors together called `vect_data` using the following code.

```{r}
vect_data <- tibble(int_vect, rand_vect, TF_vect, TF_vect2)
```

### 1.5

Coerce `rand_vect` to character class using `as.character()`. Save this vector as `rand_char_vect`. How is the output for `rand_vect` and `rand_char_vect` different?

```{r 1.5response}
rand_char_vect <- as.character(rand_vect)
rand_char_vect # Numbers now have quotation marks
```

### 1.6

Read in the National Wastewater Surveillance System (NWSS) SARS-CoV-2 Wastewater data using the url link and the code provided. 

The NWSS uses water from different sewage treatment plants to test for the SARS-CoV-2 virus, as a way to estimate how many COVID infections a community is experiencing.

```{r}
sars_ww <- 
  read_csv(file = "https://daseh.org/data/SARS-CoV-2_Wastewater_Data.csv")
```

### 1.7

Use the `mutate()` function to create a new column named `date_formatted`, based on the `date_end` column. Hint: use `mdy()` function. Reassign to `sars_ww`.

`date_end`: This is the last date of the sampling window. A sampling window is used to measure change in viral concentration.

```
# General format
NEWDATA <- OLD_DATA %>% mutate(NEW_COLUMN = OLD_COLUMN)
```

```{r 1.7response}
sars_ww <- sars_ww %>% mutate(date_formatted = mdy(date_end))
```


# Practice on Your Own!

### P.1

Move the `date_formatted` variable to be before `date_end` using the `relocate` function. Take a look at the data using `glimpse()`. Note the difference between `date_end` and `date_formatted` columns.

```
# General format
NEWDATA <- OLD_DATA %>% relocate(COLUMN1, .before = COLUMN2)
```

```{r P.1response}
sars_ww <- sars_ww %>% relocate(date_formatted, .before = date_end)

# alternative
# sars_ww <- sars_ww %>% select(date_end, date_formatted, everything()) %>% head() 

glimpse(sars_ww)

```

### P.2

Use `range()` function on `date_formatted` variable to display the range of dates in the data set. How does this compare to that of `date_end`? Why? (Hint: use the pull function first to pull the values.)

```{r P.2response}
pull(sars_ww, date_formatted) %>% range()
pull(sars_ww, date_end) %>% range()

# The max of `pull(sars_ww, date_end) %>% range()` is numerical not based on date.
```
