---
title: "Factors Lab"
output: html_document
editor_options: 
  chunk_output_type: console
---

Load all the packages we will use in this lab. 

```{r message=FALSE}
library(tidyverse)
```

### 1.0

Load the CalEnviroScreen dataset and use `select` to choose the `CaliforniaCounty`, `ImpWaterBodies`, and `ZIP` variables. Then subset this data using `filter` to include only the California counties Amador, Napa, Ventura, and San Francisco. Name this data "ces". 

`ImpWaterBodies`: measure of the number of pollutants across all impaired water bodies within a given distance of populated areas.

```{r}
ces <- 
  read_csv("https://daseh.org/data/CalEnviroScreen_data.csv") |> 
  select(CaliforniaCounty, ImpWaterBodies, ZIP) |>
  filter(CaliforniaCounty %in% c("Amador", "Napa", "Ventura", "San Francisco"))
```

### 1.1

Create a boxplot showing the difference in groundwater contamination threats (`ImpWaterBodies`) among Amador, Napa, San Francisco, and Ventura counties (`CaliforniaCounty`). **Hint**: Use `aes(x = CaliforniaCounty, y = ImpWaterBodies)` and `geom_boxplot()`.

```{r 1.1response}

```

### 1.2

Use `count` to count up the number of observations of data for each `CaliforniaCounty` group.

```{r 1.2response}

```

### 1.3

Make `CaliforniaCounty` a factor using the `mutate` and `factor` functions. Use the `levels` argument inside `factor` to reorder `CaliforniaCounty`. Reorder this variable so the order is now San Francisco, Ventura, Napa, and Amador. Assign the output the name "ces_fct".

```{r 1.3response}

```

### 1.4

Repeat question 1.1 and 1.2 using the "ces_fct" data. You should see different ordering in the plot and `count` table.

```{r 1.4response}

```


# Practice on Your Own!

### P.1
Subset `ces_fct` so that it only includes data from Ventura county. Then convert `ZIP` (zip code) into a factor using the `mutate` and `factor` functions. Do not add a `levels =` argument.

```{r P.1response}

```

### P.2

We want to create a new column that contains the group-level median values for `ImpWaterBodies`. 

- Using the "ces_Ventura" data, group the data by `ZIP` using `group_by`
- Then, use `mutate` to create a new column `med_ImpWaterBodies` that is the median of `ImpWaterBodies`. 
- **Hint**: Since you have already done `group_by`, a median `ImpWaterBodies` will automatically be created for each unique level in `ZIP`. Use the `median` function with `na.rm = TRUE`.

```{r P.2response}

```

### P.3

We want to make a plot of the `med_ImpWaterBodies`  column we created above in the `ces_Ventura`, separated by `ZIP`. Using the `forcats` package, create a plot that:

- Has `ZIP` on the x-axis
- Uses the `mapping` argument and the `fct_reorder` function to order the x-axis by `med_ImpWaterBodies`
- Has `med_ImpWaterBodies` on the y-axis
- Is a boxplot (`geom_boxplot`)
- Has the x axis label of "Zipcode"
(Don't worry if you get a warning about not being able to plot `NA` values.)

Save your plot using `ggsave()` with a width of 10 and height of 3.

Which zipcode has the largest median measure of water pollution?

```{r P.3response}

```
